---
title: "Exosome Microarray Analysis: HVE cells only"
author: "Claire Levy"
date: "January 13, 2017"
output: github_document
---



## Experiment overview

Experiments were done on two cells types: Langerhans cells and vaginal epithelial cells.

HVE cells were exposed to the following:

* High concentration of exosomes ("HighExosomes")
* Low concentration of exosomes ("LowExosomes")
* Seminal supernatant
* Media control

## The following analysis is only for HVE cells

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(lumi)
library(limma)
library(pander)
library(stringr)
library(ggplot2)
library(reshape2)

```

```{r load lumibatch data and subset , cache = TRUE}

#note that the sampleNames are NOT syntactically valid, here (they have "-")
lumibatch<-lumiR(fileName = "..\\raw_data\\FinalReport.txt",
            detectionTh = 0.05,
            annotationColumn=c('ENTREZ_GENE_ID','ACCESSION', 'SYMBOL', 'PROBE_SEQUENCE', 'PROBE_START', 'CHROMOSOME', 'PROBE_CHR_ORIENTATION', 'PROBE_COORDINATES', 'DEFINITION'))

#read in pData
pData <- read.table("..\\raw_data\\SampleInformation.txt", header = TRUE)

#make names valid in pData
corrected_pData <- pData %>%
  mutate(ArrayName = str_replace(ArrayName,"-","_"))%>%
  mutate(Donor = str_replace(Donor, "-","_"))

#make the rownames of pData the sample as the ArrayName
row.names(corrected_pData)<-corrected_pData$ArrayName


#make names valid in lumibatch
sampleNames(lumibatch)<- unique(corrected_pData$ArrayName)

#put pData into an adf
adf<-AnnotatedDataFrame(data=corrected_pData)


valid_lumibatch<-new("LumiBatch",
                     assayData = assayData(lumibatch),
                     phenoData = adf,
                     detection = detection(lumibatch),
                     controlData = controlData(lumibatch),
                     featureData = featureData(lumibatch))

#subset lumibatch for just the HVE cells
HVE_lumibatch<-valid_lumibatch[,valid_lumibatch$CellType == "HVE"]
```

Plots of non-normalized data: HVE_A2 appears as the main outlier, it is donor A with low exosomes.
```{r density plot of nonnorm data}
density(HVE_lumibatch)#number of probes for each sample that occur
#at a certain log2 intensity
```


```{r sample relations plot}

dimensional_data <- plotMDS(HVE_lumibatch)#make the plot, it invisibly returns a large object that has the actual dimensions data in matrices called x and y (one for each dimension)

#make a data frame where
d <- data.frame(Dim1 = dimensional_data$x, #Dim1 is a column containing the values from the x matrix
               Dim2 = dimensional_data$y, #Dim2 is a column containing the values from the y matrix
               ArrayName = names(dimensional_data$x))#SampleID is a column that holds the column names from the x matrix, which happen to be the microarray wells.

d <- as_data_frame(merge(d, pData(HVE_lumibatch)), by = "ArrayName")
```


```{r plot nonnorm dims}

ggplot(d, aes(x=Dim1, y = Dim2))+
  geom_point(aes(shape = Donor, color = Treatment), size = 3)
```

```{r boxplot}
#boxplot
boxplot(HVE_lumibatch)
```


```{r background correction, results = 'hide'}

#the data we got from the core had no background correction (I don't think it did anyway...) so I will do it here.
B_HVE_lumibatch<-lumiB(HVE_lumibatch, method = "bgAdjust")


# VST TRANSFORMATION 
#"Stabilizing the expression variance based on
#the bead level expression variance and mean relations"

TB_HVE_lumibatch <-lumiT (B_HVE_lumibatch, method = 'vst')
#can do this to look at a plot. plotting doesn't work if I do method = "log2"
# plotVST(TB_HVE_lumibatch)
```

Plots of quantile normalized data

```{r quantile normalization and plots}

#trying quantile normalization here (not robust spline as before)
NTB_HVE_lumibatch<-lumiN(TB_HVE_lumibatch,method="quantile")

#this seems to be fine with quantile normalization...
density(NTB_HVE_lumibatch)
```

```{r normalized MDS}

norm_dimensional_data <- plotMDS(NTB_HVE_lumibatch)#make the plot, it invisibly returns a large object that has the actual dimensions data in matrices called x and y (one for each dimension)

#make a data frame where
normd <- data.frame(Dim1 = norm_dimensional_data$x, #Dim1 is a column containing the values from the x matrix
               Dim2 = norm_dimensional_data$y, #Dim2 is a column containing the values from the y matrix
               ArrayName = names(norm_dimensional_data$x))#SampleID is a column that holds the column names from the x matrix, which happen to be the microarray wells.

normd <- as_data_frame(merge(normd, pData(NTB_HVE_lumibatch)), by = "ArrayName")


#HVE_B media is hidden under low exosomes
ggplot(normd, aes(x=Dim1, y = Dim2))+
  geom_point(aes(shape = Donor, color = Treatment), size = 3)

```




```{r quantile norm boxplot}

boxplot(NTB_HVE_lumibatch)
```



```{r RSN normalization and plot}

# NrsnTB_HVE_lumibatch<-lumiN(TB_HVE_lumibatch,method="rsn")
# #This density plot looks fine too I think.
# density(NrsnTB_HVE_lumibatch)
```


```{r rns boxplot}
# boxplot(NrsnTB_HVE_lumibatch)
```

```{r QC of normalized data, results = "hide"}

#I'm going to use the quantile normalized data for now.

QNTB_HVE_lumibatch<-lumiQ(NTB_HVE_lumibatch)

```

# Non-specific filtering

Limma suggests to keep probes that are expressed above background on at least n arrays where n is smallest number of replicates assigned to any of the treatment combinations.

We have 3 donors x 4 Treatments so I will keep probes with detection levels above background in at least 3 samples.

```{r nonspecific filtering}
#this is how the detection filtering works:
#Mark the detection pvalues (there is one per probe per sample) with a 1 if <0.05 or a 0 if >0.05
# using (detection(QNTB.complete.RAW.lumi)<0.05)

#add up the 0's and 1's across each row (i.e. for all the samples)
#using rowSums

#now you have the number of detection p values <0.05 for each probe
#for all the samples (max possible = total samples)

#now tell me which probes have a rowSum of >=3 (probes will be marked as TRUE
# or false if they do or do not have >=3 rowSum)

HVE_detectedProbes <- rowSums(detection(QNTB_HVE_lumibatch)<0.05)>= 3

#now extract just those probes that are TRUE from the lumibatch

HVE_expressedProbes_lumibatch <-QNTB_HVE_lumibatch[HVE_detectedProbes,]
```

Number of probes in data set before filtering:

```{r before filtering}
dims(QNTB_HVE_lumibatch)
```

Numbers of probes in data set after filtering:
```{r after filtering}
dims(HVE_expressedProbes_lumibatch)
```

```{r targets and design}

targets <-pData(valid_lumibatch)%>%
  filter(CellType == "HVE")%>%
  select(Donor, Treatment)

Donor <-factor(targets$Donor)

Treatment <-factor(targets$Treatment, levels = c("Media","LowExosomes","HighExosomes","SeminalSupernatant"))

                   
#When I have ~0 for the intercept, each coeff represents the avg of the samples at each level of Treatment and Donor. If I had used ~1 instead, the intercept would represent the avg of the samples for Media and the next coeff would represent the increase in the average of LowExosomes OVER Media.

HVE_design <-model.matrix(~0+Treatment+Donor)
```

```{r fit and fit contrasts and eBayes}

HVE_fit <- lmFit(HVE_expressedProbes_lumibatch, design = HVE_design)

HVE_contrasts <-makeContrasts(
  MediaVsLowExos = TreatmentLowExosomes - TreatmentMedia,
  MediaVsHighExos = TreatmentHighExosomes - TreatmentMedia,
  MediaVsSup = TreatmentSeminalSupernatant - TreatmentMedia,
  LowExosVsHighExos = (TreatmentHighExosomes - TreatmentMedia)- (TreatmentLowExosomes - TreatmentMedia), levels = HVE_design)

HVE_fit2 <-contrasts.fit (HVE_contrasts, fit = HVE_fit)


HVE_fit2 <-eBayes(HVE_fit2)

```


How many DE probes  were in each contrast?
```{r, how many probes are up and down regulated}

#method=separate is same as doing topTable for all coefs separately

results <- decideTests(HVE_fit2,method="separate", adjust.method="BH",
                      p.value=0.05, lfc=0.5)

#turn the results matrix into a data frame and make the
#probeID a real column and remove the rownames

resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing


resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))

pander(summary)
```


```{r toptable for MediaVsSup contrast}

ttMediaVsSup<- topTable(HVE_fit2, coef = "MediaVsSup", adjust.method = "BH", number=Inf, p.value=0.05, lfc=0.5)
```

## CAMERA testing

#Hallmark gene sets: top 6 results

```{r CAMERA setup}

##Expression set to use

eSet<-exprs(HVE_expressedProbes_lumibatch)

#IDs from eset: ProbeIDS from the data I'm analyzing (not the pre-determined gene set). There are some duplicates in there but I am leaving them in for now because I know they will be removed non-randomly. This means less power though I think.

identifiers = as.character(fData(HVE_expressedProbes_lumibatch)[,"ENTREZ_GENE_ID"])
```

```{r camera function}
#a function to to a camera test looping over a list of contrasts
#I have already defined the identifiers, design, index and y above

contrNames<- list("MediaVsLowExos","MediaVsHighExos","MediaVsSup","LowExosVsHighExos")

#the function arguments allow you to use different contrasts and a different index
cameraTest <- function (contrName, index){
camera(y = eSet, index = index, design = HVE_design, 
            contrast = HVE_contrasts[,contrName])#subset the HVE_contrasts matrix and get one contrast at a time.
}
```


```{r Hallmark CAMERA test}
#I'm going to do a CAMERA test using the MSigDB hallmark sets on all of my contrasts
#downloaded Rdata file 22Apr16 from http://bioinf.wehi.edu.au/software/MSigDB/
load("geneSets_for_CAMERA/human_Hallmark_v5.rdata")

#I want to use the following index that goes with the MSigDB hallmark gene sets
hallmarkIndex<-ids2indices(gene.sets = Hs.H, identifiers = identifiers)

#apply the function
hallMarkResults<-lapply(contrNames, FUN = cameraTest, index = hallmarkIndex)

#name the results dfs
names(hallMarkResults)<-c("MediaVsLowExos","MediaVsHighExos","MediaVsSup","LowExosVsHighExos")

```

Media Vs Low Exos
```{r}
pander(head(hallMarkResults[["MediaVsLowExos"]]))
```

Media Vs High Exos

```{r}
pander(head(hallMarkResults[["MediaVsHighExos"]]))


# doi: 10.1042/BCJ20160006.
```

Media Vs Sup

```{r}
pander(head(hallMarkResults[["MediaVsSup"]]))

#apparently there are a lot of reactive O2 species in semen.

# Inactivation of human immunodeficiency virus type 1 by the amine oxidase-peroxidase system. PMID: 7559947


#wikipedia says
# xenobiotic metabolism (from the Greek xenos "stranger" and biotic "related to living beings") is the set of metabolic pathways that modify the chemical structure of xenobiotics, which are compounds foreign to an organism's normal biochemistry, such any drug or poison
#results look drug related, but the "foreign substance" part makes sense in our context I guess.
```

Low Exos Vs High Exos
```{r}
pander(head(hallMarkResults[["LowExosVsHighExos"]]))
```

#GO gene sets: top 6 results

```{r GO CAMERA test}


#I'm going to do a CAMERA test using the MSigDB GO sets on all of my contrasts
#downloaded 18Jan17 from http://bioinf.wehi.edu.au/software/MSigDB/
load("geneSets_for_CAMERA/human_c5_v5p2.rdata")

#I want to use the following index that goes with the GO gene sets
GOindex<-ids2indices(gene.sets = Hs.c5, identifiers = identifiers)

#apply the function
GOresults<-lapply(contrNames, FUN = cameraTest, index = GOindex)

#name the results dfs
names(GOresults)<-c("MediaVsLowExos","MediaVsHighExos","MediaVsSup","LowExosVsHighExos")


```



Media Vs Low Exos
```{r}
pander(head(GOresults[["MediaVsLowExos"]]))
```

Media Vs High Exos

```{r}
pander(head(GOresults[["MediaVsHighExos"]]))
```

Media Vs Sup

```{r}
pander(head(GOresults[["MediaVsSup"]]))
```

Low Exos Vs High Exos
```{r}
pander(head(GOresults[["LowExosVsHighExos"]]))
```